---
title: "Exemptions Analysis"
output: 
  html_document:
    df_print: paged
    theme: sandstone
author: "Eric Langowski"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, 
                      message = FALSE, fig.width = 8, fig.height = 6)
library(tidyverse)
library(sf)
library(cmfproperty)
library(mapview)
library(ggplot2)
library(tidycensus)
library(lubridate)
library(tigris)
theme_set(theme_bw())

exemptions <- read_csv('exemptions_06_21.csv.gz')

make_table <- function(inputdata){
    DT::datatable(
      inputdata, extensions = 'Buttons', 
      rownames = FALSE,
      options = list(
        dom = 'Bfrtip', 
        buttons = c('excel', 'pdf'),
        pageLength = 20)
  )
}

fred <- read_csv('CPILFESL.csv')

inflation <- fred %>% filter(month(DATE) == 1, 
                year(DATE) >= 2006) %>%
  mutate(year = year(DATE) - 1,
         index = CPILFESL[year == 2021] / CPILFESL)

parcels <- read_sf('~/../Downloads/ccgisdata - Parcel 2021.geojson')
cook_subd <- county_subdivisions('17', '031', cb=FALSE) %>%
  select(NAME) %>%
  filter(NAME != 'County subdivisions not defined')

parcels_mini <- parcels %>% tibble() %>% 
  select(pin10, latitude, longitude) %>%
  st_as_sf(coords=c('longitude', 'latitude')) %>%
  distinct(pin10, .keep_all = TRUE)

st_crs(parcels_mini) <- st_crs(cook_subd)

parcel_sub <- parcels_mini %>%
  st_join(cook_subd) %>% tibble() %>% 
  select(-geometry)
  
```


```{r}
granted <- exemptions %>% 
  select(year, pin, total_tax_rate, starts_with('exe'), -exe_total) %>%
  pivot_longer(-c(pin, year, total_tax_rate), 
               names_to='type', values_to='eav_amount') %>%
  filter(eav_amount > 0, type != 'exe_abate') %>%
  mutate(tax_amount = total_tax_rate * eav_amount,
         type = str_remove(type, 'exe_'),
         type = if_else(str_detect(type, 'vet_',), 'veteran', type)) %>%
  group_by(year, pin, total_tax_rate, type) %>%
  summarize(eav_amount = sum(eav_amount),
            tax_amount = sum(tax_amount)) %>%
  ungroup()
```

Numerous exemptions are available in Cook County's property tax system. Most exemptions act by reducing the (taxable) equalized assessed value (EAV) and vary in amount year-to-year based on local tax rates. This report analyzes the size and number of exemptions granted from 2006 to 2021.

Exemptions are processed by the [Cook County Assessor](https://www.cookcountyassessor.com/exemptions). Exemptions are defined in [State Statute](https://tax.illinois.gov/localgovernments/property/taxrelief.html).

## An example

To illustrate how the size of an exemption changes year-to-year, below is the history of one property in Gage Park / West Lawn which has had the homestead exemption from 2006 to 2021. Notice how the value of the exemption varies year-to-year but **decreases** over time.

```{r}
granted %>% filter(pin == '19123240200000') %>%
  mutate(across(where(is.numeric), round, 3)) %>%
  make_table()
```

The tax amount, or size of the exemption, is dependent on both the EAV amount and the total tax rate for each property. From 2017 to 2021, the size of the homestead exemption fell from \$683 to \$670. When accounting for inflation, the real decrease in the size of the exemption is close to 14%.

## Exemptions & Inflation

The real amount of the homestead exemption similarly fell from \$1131 on average in 2017 to \$1014 in 2021.

```{r}
inflation_exempt <- granted %>% group_by(year, type) %>%
  summarize(median_amt = median(tax_amount),
            mean_amt = mean(tax_amount)) %>%
  left_join(inflation) %>%
  mutate(median_amt = median_amt * index,
         mean_amt = mean_amt * index)
```

```{r}
ggplot(inflation_exempt %>% filter(type == 'homeowner'), 
       aes(x=year, y=mean_amt)) +
  geom_point() +
  geom_line() +
  scale_y_continuous(labels=scales::dollar) +
  labs(x='Year', y='Average Amount', 
       title='Inflation Adjusted Size of Homestead Exemption', 
       caption = '2022 dollars')
```

## Exemptions & Tax Bill Share

Share of residential tax bills which are exempted.

```{r}
total_exempt <- exemptions %>% 
  filter(between(class, 200, 299)) %>% 
  group_by(year) %>%
  summarize(
    total_tax = sum(tax_amt_pre_exe),
    total_exempt = sum(tax_amt_exe),
    pct_exemptions = total_exempt / total_tax
  )

ggplot(total_exempt, aes(x=year, y=pct_exemptions)) +
  geom_bar(stat='identity') +
  scale_y_continuous(labels=scales::percent) +
  labs(x='Year', y='Exemption Percent', title='Share of Tax Amount which is Exempted')
```


# Most Common Exemptions

The Assessor releases information on 7 types of exemptions. Below are the detailed categories. The abatement exemption is primarily given to commercial property and is excluded from this analysis. There are multiple categories of exemptions given to veterans and these categories are combined here.

Number of exemptions granted per year

```{r}
granted %>% count(year, type) %>%
  pivot_wider(names_from = type, values_from=n, values_fill=0) %>%
  make_table()
```

## Exemptions by EAV

```{r}
pdata <- granted %>% group_by(year, type) %>%
  summarize(median_amt = median(eav_amount),
            mean_amt = mean(eav_amount))
ggplot(pdata, aes(x=year, y=median_amt, color=type)) +
  geom_line() +
  geom_point() +
  labs(x='Year', y='Median EAV Amount', 
       title='Median EAV of each Exemption per Year', color='')

ggplot(pdata, aes(x=year, y=mean_amt, color=type)) +
  geom_line() +
  geom_point() +
  labs(x='Year', y='Mean EAV Amount', 
       title='Mean EAV of each Exemption per Year', color='')
```

Total amount (millions of EAV) of exemptions per year

```{r}
granted %>% group_by(year, type) %>%
  summarize(`n` = round(sum(eav_amount)/1e6)) %>%
  pivot_wider(names_from = type, values_from=n, values_fill=0) %>%
  make_table()
```

## Exemptions by Taxable Reduction

Inflation Adjusted to 2022 dollars

```{r}
pdata <- granted %>%
  group_by(year, type) %>%
  summarize(median_amt = median(tax_amount),
            mean_amt = mean(tax_amount)) %>%
  left_join(inflation) %>%
  mutate(median_amt = median_amt * index,
         mean_amt = mean_amt * index)


ggplot(pdata, aes(x=year, y=median_amt, color=type)) +
  geom_line() +
  geom_point() +
  scale_y_continuous(labels=scales::dollar) +
  labs(x='Year', y='Median Reduction Amount', 
       title='Median Amount of each Exemption per Year', color='',
       caption='2022 dollars')

ggplot(pdata, aes(x=year, y=mean_amt, color=type)) +
  geom_line() +
  geom_point() +
  scale_y_continuous(labels=scales::dollar) +
  labs(x='Year', y='Average Reduction Amount', 
       title='Average Amount of each Exemption per Year', color='', caption='2022 dollars')
```

Total amount (millions of 2022 dollars) of exemptions per year

```{r}
granted %>% group_by(year, type) %>%
  summarize(`n` = sum(tax_amount)) %>%
  left_join(inflation) %>%
  mutate(`n` = scales::dollar(round(`n` * index / 1e6))) %>%
  select(-index, -CPILFESL, -DATE) %>%
  pivot_wider(names_from = type, values_from=n, values_fill='$0') %>%
  make_table()
```

# Geography of Exemptions

To explore how exemption amounts differ across the county, our unit of analysis will be census county subdivisions which approximate townships/large cities. There are 31 subdivision in Cook County.

```{r}
mapview(cook_subd, legend=FALSE)
```

```{r}
county_sub_amounts <- granted %>% 
  mutate(pin10 = str_sub(pin, 1, 10)) %>%
  left_join(parcel_sub) %>%
  group_by(year, NAME, type) %>%
  summarize(median_amt = median(tax_amount),
            mean_amt = mean(tax_amount),
            count = n()) %>%
  left_join(inflation %>% select(year, index)) %>%
  mutate(median_amt = round(median_amt * index),
         mean_amt = round(mean_amt * index)) %>%
  select(-index) %>%
  filter(!is.na(NAME))
```

Average size of homestead exemption in 2021.

```{r}
cook1 <- 
  cook_subd %>% left_join(
  county_sub_amounts %>% filter(type == 'homeowner', year == 2021)
)

mapview(cook1, zcol='mean_amt')

```

Absolute change in real 2022 dollars in homestead exemption from 2006 to 2021. For example, in Chicago the value of the homestead exemption fell from \$1751 in 2006 to \$670 in 2021, in 2022 dollars.

```{r}
cook2 <- 
  cook_subd %>% left_join(
  county_sub_amounts %>% filter(type == 'homeowner') %>%
  group_by(NAME) %>%
  summarize(abs_change = mean_amt[year == 2021] - mean_amt[year == 2006])
)

mapview(cook2, zcol='abs_change')
```

Share of residential properties with exemptions

```{r}
exempt_share <- exemptions %>% 
  mutate(pin10 = str_sub(pin, 1, 10)) %>%
  filter(between(class, 200, 299)) %>%
  left_join(parcel_sub) %>%
  group_by(year, NAME) %>%
  summarize(pct_with_exempt = length(pin[exe_total > 0]) / n())


cook3 <- 
  cook_subd %>% left_join(
  exempt_share %>% 
    filter(year == 2021) %>% 
    mutate(pct_with_exempt = round(pct_with_exempt, 3))
)

mapview(cook3, zcol='pct_with_exempt')
```


### Average exemption amount by subdivision, 2022 dollars.

Filter table by searching for a single jurisdiction.

```{r}
county_sub_amounts %>% select(-median_amt, -count) %>%
  pivot_wider(names_from = type, values_from = mean_amt) %>%
  make_table()
```

# Notes

Data is from [PTAXSIM](https://ccao-data-science---modeling.gitlab.io/packages/ptaxsim/index.html) and Cook County Open Data. 
