---
title: "Assessor Neighborhoods Analysis"
output: 
  html_document:
    toc: true
    toc_float: true
    df_print: paged
    theme: sandstone
author: "Eric Langowski"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, 
                      message = FALSE, fig.width = 8, fig.height = 6)
library(tidyverse)
library(sf)
library(cmfproperty)
library(mapview)
library(ggplot2)
library(DT)
library(foreach)
library(doParallel)
library(tidycensus)
library(lubridate)
theme_set(theme_bw())

neighborhoods <- read_sf('Assessor - Neighborhood Boundaries.geojson')
sales <- read_csv('~/../Downloads/Assessor_-_Parcel_Sales.csv')
assess <- read_csv('~/../Downloads/Assessor_-_Historic_Assessed_Values.csv')
parcels <- read_sf('~/../Downloads/ccgisdata - Parcel 2021.geojson')

cook_tracts <- get_acs(geography='tract',
                variables = c(capitaincome = "B19301_001",
                            totalpop = "B03002_001",
                            white_alone = "B03002_003",
                            black_alone = "B03002_004",
                            asian_alone = "B03002_006",
                            hispanic = "B03002_012"),
                state = '17', county = '031', output = 'wide',
  geometry=TRUE
) %>% select(-ends_with('M')) 

cook_tracts <- st_transform(cook_tracts, st_crs(neighborhoods))

parcels_mini <- parcels %>% tibble() %>% 
  select(pin10, latitude, longitude) %>%
  st_as_sf(coords=c('longitude', 'latitude')) %>%
  distinct(pin10, .keep_all = TRUE)

st_crs(parcels_mini) <- st_crs(neighborhoods)

targ_class <- c(
  '202', '203', '204', '205', '206', '207',
  '208', '209', '210', '211', '212', '234',
  '278', '295'
)

joined <- sales %>% filter(year >= 2018,
                 class %in% targ_class,
                 num_parcels_sale == 1) %>% 
  select(pin, year, township_code,
         class, sale_price, sale_date) %>%
  mutate(pin = str_pad(pin, 14, 'left', '0')) %>%
  left_join(
    assess %>% select(pin, tax_year, class, certified_tot, mailed_tot),
    by=c('pin', 'year'='tax_year', 'class')
  ) %>%
  mutate(pin10 = str_sub(pin, 1, 10)) %>%
  left_join(parcels_mini) %>% 
  st_as_sf() %>% 
  st_join(neighborhoods %>% select(-township_code))

ratios <- reformat_data(joined %>% tibble(), 'sale_price', 'mailed_tot', 'year', TRUE) %>% tibble()

# make_report(ratios %>% filter(triad_name == 'City'), 'Cook County (mailed, City Tri)', diagnostic = TRUE)

nbhd_parcel_cnts <- assess %>% filter(tax_year == 2022,
                                      class %in% targ_class) %>%
  mutate(pin10 = str_sub(pin, 1, 10)) %>%
  left_join(parcels_mini) %>% 
  st_as_sf() %>% 
  st_join(neighborhoods %>% select(-township_code, -township_name)) %>%
  tibble() %>%
  count(town_nbhd)
```

This report analyzes the 2021 city triennial reassessment and resulting 2022 assessments for single and multifamily homes across Assessor Neighborhoods to contextualize assessment performance in Pilsen. The Cook County Assessor issues and utilizes [860 neighborhood boundaries](https://datacatalog.cookcountyil.gov/Property-Taxation/Assessor-Neighborhood-Boundaries/pcdw-pxtg) for the purpose of assessing properties.

Below is an interactive map of the Assessor Neighborhoods. These neighborhoods are "intended to represent homogeneous housing sub-markets." The neighborhoods used by the Assessor are a key step in the assessment process and often the most important property characteristic in creating valuations especially since the Assessor often has incomplete information on property conditions and renovations.

```{r}
mapview(neighborhoods, zcol='town_nbhd', legend=FALSE)
```

# Analysis of Assessments for Assessor Neighborhoods

The standard approach for evaluating the quality and fairness of assessments is through a sales ratio study. A property’s sales ratio is defined as the assessed value divided by the sale price. A sales ratio study evaluates the extent of regressivity in a jurisdiction, along with other aspects of assessment performance, by studying sales ratios for properties that sold within a specific time period. A system in which less expensive homes are systematically assessed at higher sales ratios than more expensive homes is regressive.

Neighborhoods which include more than 150 sales of single family and multi family properties between 2020 and 2022 are included in this analysis (average of 50 sales per year). 284 Neighborhoods meet this criteria which contain 73.4% of single and multi family properties in the county.

```{r}
#800727 / 1090592
# nbhd_parcel_cnts %>% filter(town_nbhd %in% final_neighborhoods$town_nbhd) %>%
#   summarize(n = sum(n))

final_neighborhoods <- neighborhoods %>% filter(
  town_nbhd %in% (ratios %>% filter(SALE_YEAR >= 2020) %>% count(town_nbhd) %>% filter(n >= 150) %>%
  pull(town_nbhd))
) 

mapview(final_neighborhoods, zcol='town_nbhd', legend=FALSE)
```

## Sales Ratio Studies

We conducted a sales ratio study for each of the target Neighborhoods. We use the Assessor mailed values here (before any appeals).

```{r}
targ_ratios <- ratios %>% filter(town_nbhd %in% final_neighborhoods$town_nbhd)

my.cluster <- parallel::makeCluster(
  16, 
  type = "PSOCK"
  )

doParallel::registerDoParallel(cl = my.cluster)


our_neigh <- targ_ratios %>% distinct(town_nbhd) %>% pull(town_nbhd)
tr_mini <- targ_ratios %>% select(SALE_YEAR, SALE_PRICE, SALE_PRICE_ADJ, ASSESSED_VALUE,
                                  ASSESSED_VALUE_ADJ, town_nbhd, RATIO) %>%
  filter(!(town_nbhd=='24031' & SALE_YEAR==2022))

output <- foreach( i=our_neigh, 
                   .packages=c('tidyverse', 'cmfproperty')) %dopar% {
  mini <- tr_mini %>% filter(town_nbhd == {i})
  rslt <- calc_iaao_stats(mini) %>% mutate(town_nbhd = {i})
  rslt
}

parallel::stopCluster(cl = my.cluster)

output2 <- map_dfr(output, bind_rows) %>% 
  tibble() %>% mutate(cod = if_else(COD <= 15, 'met', 'did not meet'),
                   prd = if_else(between(PRD, 0.98, 1.03), 'met', 'did not meet'),
                   prb = if_else(between(PRB, -0.05, 0.05), 'met', 'did not meet')) %>%
  group_by(Year)
```

### Coefficient of Dispersion (COD)

The COD is a measure of assessment uniformity, or horizontal equity. It is the average absolute percentage difference from the median sales ratio. For instance, a COD of 10 means that properties have ratios that on average deviate by 10 percent from the median ratio. The IAAO specifies that the acceptable range for COD is below 15.

```{r}
output2 %>%
  count(cod) %>%
  pivot_wider(names_from=cod, values_from=n) %>%
  mutate(percent_met = scales::percent(met / (met + `did not meet`)))
```

### Price-Related Differential (PRD)

The PRD is a measure of regressivity, or vertical equity. A PRD of 1 indicates that homes are assessed at the same rate regardless of their sale price. A PRD greater than 1 indicates that less expensive homes are assessed at higher rates than more expensive homes, while a PRD less than 1 represents the opposite situation. The IAAO specifies that the acceptable range of PRD is .98 to 1.03.

```{r}
output2 %>%
  count(prd) %>%
  pivot_wider(names_from=prd, values_from=n) %>%
  mutate(percent_met = scales::percent(met / (met + `did not meet`)))
```

## Coefficient of Price-Related Bias (PRB)

The PRB is another quantitative measure of regressivity (vertical equity) which is an alternative to the PRD. PRB is a measure of how much assessed values change as a property’s market value increases. The IAAO specifies that the acceptable range for PRB is between -0.05 and 0.05

```{r}
output2 %>%
  count(prb) %>%
  pivot_wider(names_from=prb, values_from=n) %>%
  mutate(percent_met = scales::percent(met / (met + `did not meet`)))
```

## Rate of Assessment

The relationship between assessments and sale prices is regressive if less valuable homes are assessed at higher rates (relative to the value of the home) than more valuable homes. To evaluate regressivity in assessments, the figure below presents a binned scatter plot of sales ratios against sale prices.

For this graph, property sales have been sorted into deciles for all target neighborhoods (10 bins of equal size based on sale price), each representing 10% of all properties sold. Each dot represents the average sale price and average sales ratio for each respective decile of properties. This graph compares the most recent values for 2022 (solid line) with the average across all years of observation from 2018 to 2022 (dashed line). All values were adjusted for inflation to 2022 dollars to facilitate comparisons.

If sale prices are a fair indication of market value and if assessments were fair and accurate, the would be a flat line indicating that sales ratios do not vary systematically according to sale price. A downward sloping line indicates that less expensive homes are over-assessed compared to more expensive homes and is evidence of regressivity.

```{r}
binned <- 
  cmfproperty::binned_scatter(
    targ_ratios,
    min_reporting_yr = 2018,
    max_reporting_yr = 2022,
    jurisdiction_name = "Cook County, IL"
  )
```

`r binned[[1]]`

```{r}
binned[[2]]
```

## Intra-Neighborhood Sales Ratios

A key way to view regressivity within neighborhoods is by comparing the average assessment rate of the highest valued homes to the lowest valued homes. Here we will calculate the average rate of assessment of the bottom 20% and top 20% of sold homes in 2022 for each neighborhood. Below is presented the average levels of assessment.

Countywide, the bottom 20% of home sales in each neighborhood are assessed at higher rates than the top 20% of sales.

```{r}
bin_means <- targ_ratios %>% filter(SALE_YEAR == 2022) %>%
  group_by(town_nbhd) %>%
  mutate(bin = ntile(SALE_PRICE, 5)) %>%
  group_by(town_nbhd, bin) %>%
  summarize(mean_ratio = mean(RATIO),
            mean_sp = mean(SALE_PRICE)) %>%
  pivot_wider(names_from=bin, values_from=c(mean_ratio, mean_sp)) %>%
  mutate(ratio_1_5 = mean_ratio_1 / mean_ratio_5)

bm_long <- bin_means %>% select(bottom_20=mean_ratio_1, top_20=mean_ratio_5, ratio_1_5) %>%
         left_join(neighborhoods %>% tibble() %>% select(triad_name, town_nbhd)) %>%
  pivot_longer(!c(town_nbhd, triad_name))

ggplot(bm_long %>% filter(name != 'ratio_1_5'), aes(x=value, fill=name)) +
  geom_histogram() +
  geom_rug(alpha=0.2) +
  labs(x='Average Ratio', y='Count', title='Average Assessment Rates by Neighborhood', 
       caption='Compares the rate of assessment in the bottom and top 20% of homes by triad (2022)',
       fill='') +
  facet_wrap(~c(triad_name)) +
  theme(legend.position='bottom')
```

This relationship is also visualized by calculating the ratio between the assessment rate of the bottom 20% and top 20% of home sales. If high and low valued properties were assessed at the same rate the ratio would be 1. Regressivity ranges from the least regressive neighborhood, 18011 near Elgin, where low valued properties are assessed at rates only 5% higher than high valued properties to the most regressive neighborhood, 12162 near Chicago Heights, where low valued properties are assessed at rates 358% higher.

```{r}
ggplot(bm_long %>% filter(name == 'ratio_1_5'), aes(x=value)) +
  geom_histogram() +
  geom_rug(alpha=0.2) +
  labs(y='Count', x='Ratio of Assessment', title='Regressivity is Present on the Neighborhood level',
       caption='Shown is the ratio of assessment rate between the bottom and top 20% of homes.') +
  facet_wrap(~triad_name)
```

```{r}
neighborhood_bins <- targ_ratios %>%
  filter(SALE_YEAR > 2018) %>%
  group_by(town_nbhd, SALE_YEAR) %>%
  mutate(bin = ntile(SALE_PRICE, 5)) %>%
  group_by(SALE_YEAR, bin) %>%
  summarize(mean_ratio = mean(RATIO),
            mean_sp = mean(SALE_PRICE))

ggplot(neighborhood_bins, aes(x=mean_sp/1e3, y=mean_ratio)) +
  geom_point() +
  geom_line() +
  scale_x_continuous(labels=scales::dollar) +
  facet_wrap(~SALE_YEAR) +
  labs(x='Average Sale Price ($ Thosuands)', y='Average Ratio',
       title='Assessments are Highly Regressive within Neighborhood',
       caption='In 2020, the 20% of lowest valued homes per neighborhood \nwere assessed on average 57% higher than the highest 20%.'
       )
```

While assessment quality on the township and county level has shown significant improvement, intra-neighborhood assessment quality is much lower. Many neighborhoods have relatively low numbers of property sales per year to conclusively evaluate assessments. Still, from 2019 to 2022 across the 163 neighborhoods we evaluated, none of the neighborhoods had a single year in which the 20% of lowest valued properties were assessed at less than or the same rate as the highest valued properties. On average, these lowest valued properties were assessed 56% higher in 2020, 47% higher in 2021, and 52% higher in 2022 than the highest valued properties **in the same Assessor Neighborhood.**

## 2022 Pilsen Assessments

The following figures compare assessment valuation metrics among the analysis neighborhoods to Pilsen. Assessments in Pilsen are more regressive and less uniform than comparable neighborhoods meeting only the IAAO standard for PRD. This information was not available to the Assessor at the time assessments were created.

```{r}
pilsen22 <- output2 %>% filter(town_nbhd == 77141, Year == 2022)

ggplot(output2 %>% filter(Year == 2022)) + 
  geom_histogram(aes(COD)) + 
  geom_rug(aes(COD), alpha = 0.2) +
  geom_vline(size=2, aes(xintercept = pilsen22 %>% pull(COD)), color="blue") +
  labs(x='COD', y='Count', title='2022 COD Neighborhood Distributions', subtitle='Pilsen (blue line)')
```

```{r}
ggplot(output2 %>% filter(Year == 2022)) + 
  geom_histogram(aes(PRD)) + 
  geom_rug(aes(PRD), alpha = 0.2) +
  geom_vline(size=2, aes(xintercept = pilsen22 %>% pull(PRD)), color="blue") +
  labs(x='PRD', y='Count', title='2022 PRD Neighborhood Distributions', subtitle='Pilsen (blue line)')
```

```{r}
ggplot(output2 %>% filter(Year == 2022)) + 
  geom_histogram(aes(PRB)) + 
  geom_rug(aes(PRB), alpha = 0.2) +
  geom_vline(size=2, aes(xintercept = pilsen22 %>% pull(PRB)), color="blue") +
  labs(x='PRB', y='Count', title='2022 PRB Neighborhood Distributions', subtitle='Pilsen (blue line)')
```

```{r}
pilsen_r15 <- bm_long %>% filter(name == 'ratio_1_5', 
                            town_nbhd == 77141) %>% 
  pull(value)

ggplot(bm_long %>% filter(name == 'ratio_1_5')) + 
  geom_histogram(aes(value)) + 
  geom_rug(aes(value), alpha = 0.2) +
  geom_vline(size=2, aes(xintercept = pilsen_r15), color="blue") +
  labs(x='Ratio of Assessment', 
       y='Count', 
       title='2022 Ratio of Assessment Neighborhood Distributions', 
       subtitle='Pilsen (blue line)')

# ecdf(bm_long %>% filter(name == 'ratio_1_5', triad_name == 'North') %>% pull(value))(pilsen_r15)
# ecdf(bm_long %>% filter(name == 'ratio_1_5', triad_name == 'South') %>% pull(value))(pilsen_r15)
# ecdf(bm_long %>% filter(name == 'ratio_1_5', triad_name == 'City') %>% pull(value))(pilsen_r15)
```

Pilsen's ratio of assessment is greater than all neighborhoods analyzed in the North Triad, 85% of the City Triad, and 77% of the South Triad. One reason for the observed regressivity could be that the Assessor Neighborhood includes properties which are distinctly different from each other---either in property types/characteristics or in housing submarkets. Neighborhood 77141 includes 18th street from the Chicago River to Western Avenue.

# Appendix

## Small Neighborhoods

Many Neighborhoods have few single/multi family properties. Neighborhoods below have fewer than 500 properties and will be excluded.

```{r}
small_neighborhoods <- neighborhoods %>% filter(town_nbhd %in%
                                   (nbhd_parcel_cnts %>% filter(n < 500) %>% pull(town_nbhd)))

mapview(small_neighborhoods, zcol='town_nbhd', legend=FALSE)
```

## Largest Neighborhoods

Some Neighborhoods have many single/multi family properties. The top fifty Neighborhoods are shown below (about 4000 or more properties in each).

```{r}
large_neighborhoods <- neighborhoods %>% filter(town_nbhd %in%
                                   (nbhd_parcel_cnts %>% slice_max(order_by=`n`, n=50) %>% pull(town_nbhd)))

mapview(large_neighborhoods, zcol='town_nbhd', legend=FALSE)
```


## Data Notes

This report uses the following sources:

- [Assessor - Neighborhood Boundaries](https://datacatalog.cookcountyil.gov/Property-Taxation/Assessor-Neighborhood-Boundaries/pcdw-pxtg)
- [Assessor - Parcel Sales](https://datacatalog.cookcountyil.gov/Property-Taxation/Assessor-Parcel-Sales/wvhk-k5uv)
- [Assessor - Historic Assessed Values](https://datacatalog.cookcountyil.gov/Property-Taxation/Assessor-Historic-Assessed-Values/uzyt-m557)
- [ccgisdata - Parcel 2021](https://datacatalog.cookcountyil.gov/Property-Taxation/ccgisdata-Parcel-2021/77tz-riq7)

An important caveat is that the Assessor does not currently release information on which sales they have determined to be arm's length transactions. In place of using that information, the standard arm's length ratio filtering methodology is used.