---
title: "Pilsen Neighborhood Boundaries Analysis"
output: 
  html_document:
    toc: true
    toc_float: true
    df_print: paged
    theme: sandstone
author: "Eric Langowski"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, 
                      message = FALSE, fig.width = 8, fig.height = 6)
library(tidyverse)
library(sf)
library(cmfproperty)
library(mapview)
library(ggplot2)
library(DT)
library(foreach)
library(doParallel)
library(tidycensus)
library(lubridate)
library(ggmap)
theme_set(theme_bw())

neighborhoods <- read_sf('Assessor - Neighborhood Boundaries.geojson')
sales <- read_csv('~/../Downloads/Assessor_-_Parcel_Sales.csv')
assess <- read_csv('~/../Downloads/Assessor_-_Historic_Assessed_Values.csv')
parcels <- read_sf('~/../Downloads/ccgisdata - Parcel 2021.geojson')

cook_tracts <- get_acs(geography='tract',
                variables = c(capitaincome = "B19301_001",
                            totalpop = "B03002_001",
                            white_alone = "B03002_003",
                            black_alone = "B03002_004",
                            asian_alone = "B03002_006",
                            hispanic = "B03002_012"),
                state = '17', county = '031', output = 'wide',
  geometry=TRUE
) %>% select(-ends_with('M')) 

cook_tracts <- st_transform(cook_tracts, st_crs(neighborhoods))

parcels_mini <- parcels %>% tibble() %>% 
  select(pin10, latitude, longitude) %>%
  st_as_sf(coords=c('longitude', 'latitude')) %>%
  distinct(pin10, .keep_all = TRUE)

st_crs(parcels_mini) <- st_crs(neighborhoods)

targ_class <- c(
  '202', '203', '204', '205', '206', '207',
  '208', '209', '210', '211', '212', '234',
  '278', '295'
)

joined <- sales %>% filter(year >= 2018,
                 class %in% targ_class,
                 num_parcels_sale == 1) %>% 
  select(pin, year, township_code,
         class, sale_price, sale_date) %>%
  mutate(pin = str_pad(pin, 14, 'left', '0')) %>%
  left_join(
    assess %>% select(pin, tax_year, class, certified_tot, mailed_tot),
    by=c('pin', 'year'='tax_year', 'class')
  ) %>%
  mutate(pin10 = str_sub(pin, 1, 10)) %>%
  left_join(parcels_mini) %>% 
  st_as_sf() %>% 
  st_join(neighborhoods %>% select(-township_code))

ratios <- reformat_data(joined %>% tibble(), 'sale_price', 'mailed_tot', 'year', TRUE) %>% 
  tibble() %>%
  filter(town_nbhd == '77141')

nbhd_parcel_cnts <- assess %>% filter(tax_year == 2022,
                                      class %in% targ_class) %>%
  mutate(pin10 = str_sub(pin, 1, 10)) %>%
  left_join(parcels_mini) %>% 
  st_as_sf() %>% 
  st_join(neighborhoods %>% select(-township_code, -township_name)) %>%
  tibble() %>%
  count(town_nbhd)

my_pal <- c("#4B0055", "#009B95", "#FDE333")

```


```{r}
pilsen_tracts <- neighborhoods %>% filter(town_nbhd %in% c(77141)) %>%
  st_intersection(cook_tracts) %>%
  select(GEOID:hispanicE, town_nbhd) %>%
  mutate(group = case_when(
    GEOID %in% c('17031841200', '17031841300', '17031842900') ~ 'West',
    GEOID %in% c('17031310900', '17031310800', '17031310700') ~ 'Central',
    GEOID %in% c('17031310400', '17031310300', '17031310200', '17031310500', '17031310600', '17031843200') ~ 'East'
  ))

neigh_assess <- assess %>% filter(tax_year >= 2019, 
                                  class %in% targ_class) %>%
  mutate(pin10 = str_sub(pin, 1, 10)) %>%
  left_join(parcels_mini %>% 
              st_join(neighborhoods %>% select(-township_code, -township_name))
            ) %>% 
  filter(town_nbhd == 77141) %>%
  st_as_sf() %>%
  st_join(pilsen_tracts %>% select(-town_nbhd)) 

tract_av <- neigh_assess %>% 
  tibble() %>%
  group_by(tax_year, GEOID) %>%
  summarize(avg_av = round(mean(certified_tot)))  %>%
  pivot_wider(names_from = tax_year, values_from=avg_av) %>%
  mutate(pct_change = round((`2021` - `2020`) / `2020`, 3))

pilsen_ratios <- ratios %>% 
  filter(town_nbhd %in% c(77141),
         SALE_YEAR <= 2022) %>%
  st_as_sf() %>%
  st_join(pilsen_tracts %>% select(-town_nbhd)) %>%
  mutate(
    lon = st_coordinates(.)[,1],
    lat = st_coordinates(.)[,2],
    group = case_when(
      group == 'East' & lon <= -87.6637 ~ 'Central',
      group == 'West' & lon >= -87.671 ~ 'Central',
      TRUE ~ group),
    sale_date = mdy(sale_date),
         quarter = floor_date(sale_date, 'quarter')) #sf

targ <- pilsen_ratios %>% tibble() %>% #tibble, not sf
          filter(between(SALE_YEAR, 2019, 2020))

pilsen_tracts <- pilsen_tracts %>% 
  filter(GEOID %in% (pilsen_ratios %>% distinct(GEOID) %>% pull(GEOID))) %>%
  left_join(targ %>% group_by(GEOID) %>% summarize(Count = n(), `Avg. SP` = round(mean(SALE_PRICE), 0),
                                                   `Avg. Ratio` = round(mean(RATIO), 3))) %>%
  left_join(tract_av)
```

Pilsen is broadly within a single Assessor Neighborhood, 77141. Two other neighborhoods (77131 and 77132) overlap slightly with Pilsen but do not include a significant number of residential properties. Neighborhood 77115 (South Lawndale/Little Village) is also included for comparison.

```{r}
p_area <- c('77141', '77131', '77132', '77115')

p_neigh <- neighborhoods %>%
          filter(town_nbhd %in% p_area)

mapview(p_neigh, zcol='town_nbhd', alpha=0)
```

## Trends

Census tract 17031842900 is excluded from this section as it contains a small number of residential properties. A series of maps follows which shows trends by census tract in Pilsen. In 2019 and 2020, observe that central Pilsen has the highest average sales and west Pilsen has the lowest average ratios/rate of assessment. This then corresponds to west Pilsen receiving the highest average increase in assessed values in 2021 (81%).

Below is a map by tract which shows the average sale price of residential properties in 2019 and 2020.

```{r}
pt <- pilsen_tracts %>% filter(GEOID != '17031842900')

mapview(pt, zcol='Avg. SP', legend=TRUE, alpha.regions=0.75)
```

Below is a map by tract which shows the average sale ratio for 2019 and 2020.

```{r}
mapview(pt, zcol='Avg. Ratio', legend=TRUE, alpha.regions=0.75)
```


Below is a map by tract which shows the average percentage increase in assessed value by tract between 2020 and 2021.

```{r}
mapview(pt, zcol='pct_change', legend=TRUE, alpha.regions=0.75)
```



# Analysis Sub-Neighborhoods

Utilizing Census Tracts within Neighborhood 77141, Pilsen properties were split into sub-neighborhoods. Below is a map of census tracts and the sub-neighborhoods. Each of these groups represents about 1/3rd of the recent sales in Pilsen.

```{r}
mapview(pilsen_tracts, zcol='GEOID', legend=FALSE, alpha.regions=0.2) +
  mapview(pilsen_ratios, zcol='group', color=my_pal)
```

Sales per group per year. These groups have a small number of sales per year which is a limitation of this analyses.

```{r}
targ %>%
  group_by(group, SALE_YEAR) %>%
  summarize(count = n())  %>%
  pivot_wider(names_from=SALE_YEAR, values_from=count, values_fill=0)
```

The following graphs use sales from 2018 to 2022 to demonstrate market trends across the three groups.

```{r}
sale_trends <- pilsen_ratios %>% tibble() %>% group_by(group, SALE_YEAR) %>%
  summarize(`Mean SP` = mean(SALE_PRICE))

ggplot(pilsen_ratios %>% tibble(), aes(x=sale_date, y=`SALE_PRICE`)) +
  geom_smooth() +
  geom_vline(xintercept = ymd('2021-01-01')) +
  scale_y_continuous(labels=scales::dollar) +
  scale_x_date(date_labels = '%m/%y') +
  labs(x='Sale Date', y='Average Sale Price', 
       title='Sale Price Trends', caption='January 2021 (vertical line)') +
  facet_wrap(~group)


pr_bin <- pilsen_ratios %>% tibble() %>%
  filter(SALE_YEAR > 2018) %>%
  group_by(group, SALE_YEAR) %>%
  mutate(bin = ntile(SALE_PRICE, 4)) %>%
  group_by(group, SALE_YEAR, bin) %>%
  summarize(
    mean_ratio = mean(RATIO),
    mean_sp = mean(SALE_PRICE),
    max_ratio = max(RATIO),
    min_ratio = min(RATIO)
  )
```

The following graph shows information on the average ratio by quartile of sale price for each group and year. Note that in 2019 west Pilsen is assessed on average below central and east Pilsen but in 2022 it is assessed at the highest rate.

```{r}
ggplot(pr_bin, aes(x=mean_sp, y=mean_ratio, color=group)) +
  geom_point(size=1.5) +
  geom_line(size=1.5) +
  facet_wrap(~SALE_YEAR) +
  scale_color_manual(values=my_pal) +
  labs(x='Quartile Average Sale Price', y='Average Ratio', 
       title='Average Sale Price by Year and Quartile') +
  scale_x_continuous(labels=scales::dollar)
```

## Alternate Sales Ratio Study

In order to determine if the Assessor should have used smaller delineations of Pilsen for the 2021 reassessment, we can only utilize information available for that reassessment (2019 and 2020 sales). A secondary analysis of the 2021 and 2022 sales against the 2021 reassessments is also included to evaluate the performance of the three proposed subgroups. The first group is shown as the 19/20 study and the second group is shown as the 21/22 study on the following graphs.

```{r, include=FALSE}
mod_ratios <- pilsen_ratios %>% 
  tibble() %>%
  mutate(group = 'All') %>%
  bind_rows(pilsen_ratios %>% tibble()) %>%
  filter(between(SALE_YEAR, 2019, 2022)) %>%
  mutate(study = if_else(SALE_YEAR == 2019 | SALE_YEAR == 2020, '19/20', '21/22'),
         SALE_YEAR = 2021) %>%
  select(-c(certified_tot, ASSESSED_VALUE, TAX_YEAR, RATIO, arms_length_transaction, 
            SALE_PRICE_ADJ, ASSESSED_VALUE_ADJ)) %>%
  left_join(
    assess %>% select(pin, tax_year, class, certified_tot, mailed_tot),
    by=c('pin', 'SALE_YEAR'='tax_year', 'class')
  ) %>%
  reformat_data('SALE_PRICE', 'mailed_tot', 'SALE_YEAR', FALSE) %>%
  mutate(g_study = paste0(group, "_", study)) %>%
  filter(!is.na(RATIO))
  

output <- tibble()
for(neigh in mod_ratios %>% distinct(g_study) %>% pull(g_study)){
  mini <- mod_ratios %>% filter(g_study == {neigh})
  output <- bind_rows(output, calc_iaao_stats(mini) %>% mutate(g_study = {neigh}))
}

output <- output %>% mutate(
  group = str_sub(g_study, 1, str_length(g_study) - 6),
  study = str_sub(g_study, str_length(g_study) - 4, str_length(g_study)),
)
```


```{r}
ggplot(output, aes(x=group, y=COD, fill=study)) +
  geom_bar(position='dodge', stat='identity') +
  labs(title='COD By Year and Group') 

ggplot(output, aes(x=group, y=PRB, fill=study)) +
  geom_bar(position='dodge', stat='identity') +
  labs(title='PRB By Year and Group') 

output %>% select(study, group, PRD)
```


## Demographic Information

Census tract 17031843200 is excluded from this section as it contains a small number of residential properties.

```{r}
pilsen_demos <- pilsen_tracts %>% 
  filter(!(GEOID %in% c('17031842900'))) %>%
  tibble() %>%
  mutate(income_person = capitaincomeE * totalpopE) %>%
  select(group, income_person, totalpopE, 
         white_aloneE, black_aloneE, 
         asian_aloneE, hispanicE) %>%
  group_by(group) %>%
  summarize(across(everything(), sum)) %>%
  mutate(per_capita = scales::dollar(round(income_person / totalpopE)),
         per_hisp = 
           scales::percent(hispanicE / totalpopE))


med_ratio <- targ %>% filter(SALE_YEAR == 2019 | SALE_YEAR == 2020) %>%
  summarize(med=median(RATIO)) %>% pull(med)

pilsen_summary <- pilsen_ratios %>% tibble() %>%
  mutate(study = if_else(SALE_YEAR == 2019 | SALE_YEAR == 2020, '19/20', '21/22')) %>%
  group_by(study, group) %>%
  summarize(avg_ratio = round(mean(RATIO), 3),
            avg_sp = scales::dollar(mean(SALE_PRICE)),
            #med_ratio = round(median(RATIO), 3),
            sales = n())

summary_info <- pilsen_demos %>% 
  select(group, `Total Pop.` = totalpopE, 
         `Hispanic Pop.` = hispanicE,
         `Pct. Hispanic` = per_hisp,
         `Per Capita Income` = per_capita) %>%
  left_join(pilsen_summary)

DT::datatable(summary_info %>% select(study, group, everything()) %>% arrange((study)),
              extensions = 'Buttons', options = list(
    dom = 'Bfrtip',
    buttons = c('csv', 'excel', 'pdf')
  )
)
```

# Appendix

## Changes in AV

Average Assessed Value by tract and percent change between 2020 and 2021.

```{r}
DT::datatable(tract_av,
              extensions = 'Buttons', options = list(
    dom = 'Bfrtip',
    buttons = c('csv', 'excel', 'pdf')
  )
)
```


## Pilsen Residential Parcels

```{r}
pilsen_parcels <- parcels %>% 
  filter(assessornbhd == 141, politicaltownship == 'Town of West',
         assessorbldgclass %in% targ_class)

mapview(pilsen_parcels)
```

## Data Export

Pilsen Sales (77141)

```{r}
DT::datatable(pilsen_ratios %>% tibble() %>%
  select(pin, SALE_YEAR, class, SALE_PRICE, ASSESSED_VALUE),
              extensions = 'Buttons', options = list(
    dom = 'Bfrtip',
    buttons = c('csv', 'excel', 'pdf')
  )
)
```


## Data Notes

This report uses the following sources:

- [Assessor - Neighborhood Boundaries](https://datacatalog.cookcountyil.gov/Property-Taxation/Assessor-Neighborhood-Boundaries/pcdw-pxtg)
- [Assessor - Parcel Sales](https://datacatalog.cookcountyil.gov/Property-Taxation/Assessor-Parcel-Sales/wvhk-k5uv)
- [Assessor - Historic Assessed Values](https://datacatalog.cookcountyil.gov/Property-Taxation/Assessor-Historic-Assessed-Values/uzyt-m557)
- [ccgisdata - Parcel 2021](https://datacatalog.cookcountyil.gov/Property-Taxation/ccgisdata-Parcel-2021/77tz-riq7)

The Assessor does not currently release information on which sales they used to value homes (arm's length transactions). The IAAO arm's length standard was applied here.

In all cases, the Assessor mailed assessment value is used (assessments before any appeals). This would then *not* include the COVID-19 adjustments made by the Assessor in May 2020 which occurred to certified assessor values in tax year 2020.
